\documentclass[11pt]{article}

\input{../common/common-defs}
\usepackage{graphicx}
\usepackage{../common/code}

\title{Inline BOM}
\author{The Manticore Group}
\date{Draft of \today}

\begin{document}
\maketitle

\section{Overview}
Inline BOM is a language extension that allows us to embed BOM code in PML programs. It is the home of our \emph{scheduling language}, in which we build the basis for our high-level parallel constructs\cite{manticore-sched-icfp08}. We also use inline BOM to implement a variety of other language mechanisms, including some of the following.
\begin{itemize}
\item process schedulers
\item parallel constructs, \eg{}, futures, \texttt{pval}, \etc{}
\item primitive operations, \eg{}, operations over 32-bit ints
\item foreign calls
\item synchronization / concurrent data structures, \ie{}, implementing CML
\end{itemize}

\subsection{Rationale}
One might reasonably as why we use inline BOM. Our primary motivation is to separate unsafe language features out of PML. Doing so gives PML a cleaner semantics, thereby simplifying the development model and enablinig more aggressive optimizations. We can use inline BOM as a sandbox for our unsafe code.

Inline BOM includes the following list of features not found in PML.
\begin{itemize}
\item scheduling primitives (\texttt{run}, \texttt{forward}, \etc{})
\item atomic operations, \eg{} compare-and-swap
\item precise byte layouts of objects
\item mutable memory
\item control over whether an object gets allocated in the local or global heap
\item C calls
\item primitive operations
\end{itemize}

\section{Design}
We have designed inline BOM as a conservative extension of our module language. Below we describe these forms and their semantics.

\subsection{Inline BOM declarations}
Primcode declarations allow us to make BOM declarations, which are types, C function prototypes, or HLOps definitions. Primcode declarations occur in top-level declarations of modules. They obey the same scoping rules as other top-level declarations.
\begin{centercode}
  _primcode (
    BOMDecl ...
  )
\end{centercode}

\subsubsection{Example: BOM type declaration}
We can define the state field of single-toucher futures as follows. The state can be either a non-pointer flag, or a pointer to either the value of the future or the continuation of a blocked thread. 
\begin{centercode}
  _primcode( 
    (*
     * a future_state word contains one of the following values:
     *          EMPTY_F
     *          STOLEN_F
     *          EVAL_F
     *          FULL      value
     *          WAITING   cont
     *)
    typedef future1_state = any; 
  )
\end{centercode}

\subsubsection{Example: C function prototype}
The following snippet contains a prototype for a print function, which is defined in some external C file.
\begin{centercode}
  _primcode(
    extern void M_Print(void*)
  )
\end{centercode}

\subsubsection{Example: HLOps}
The following HLOp prints its parameter string.
\begin{centercode}
  _primcode(
    define inline @print-ln(msg : string / exh : exh) : () =
      do ccall M_Print(msg)
      return ()
  )
\end{centercode}

\subsection{Types}
We can make type declarations that are visible to PML and BOM code. In PML, the type t is an abstract type, but in BOM, it has the type bomTy.
\begin{centercode}
  type ('a, ...) t = _prim(bomTy)
\end{centercode}

\subsubsection{Example}
We can define the type of single-toucher futures as follows. The future structure is just a mutable pair of the state and the thunk.
\begin{centercode}
  type future1 = _prim(![future1_state, thunk])
\end{centercode}

\subsection{Exporting BOM definitions}
We can bind BOM functions and HLOps to PML identifiers. Because the translation from BOM types to PML types is undefined, we require the programmer to ascribe the bound variable a PML type. 
\begin{centercode}
  val f : ty = _prim(bomId)
\end{centercode}

\subsubsection{Example}
We export the HLOp \texttt{@f} to PML as the function \texttt{f}.
\begin{centercode}
  _primcode(
    define @f(_ : unit / exh : exh) : unit = ...;
  )
  val f : unit -> unit = _prim(@f)
\end{centercode}

\subsection{Importing PML datatypes}
BOM code can refer to datatypes and constructors defined in PML.

\subsubsection{Example}
The HLOp \texttt{mk-x} creates an element of the datatype \texttt{t}.
\begin{centercode}
  datatype t = X of int | Y
  
  _primcode(
    define @mk-x(x : int / exh : exh) : t =
      return(X(x))
  )
\end{centercode}

\subsection{Importing PML declarations}
BOM code can refer to PML variables that are bound at the type level.
\begin{centercode}
  let f : bomTy = pmlvar(pmlId)
\end{centercode}

\subsubsection{Example}
Here we bind the list application function for use in BOM.
\begin{centercode}
  let app : fun( [fun(any / exh -> unit), List.list] / exh) = 
                     pmlvar(List.app)
\end{centercode}

\bibliographystyle{alpha}
\bibliography{../common/manticore}

\end{document}
